---
title: "ghcn"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ghcn}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ghcn)
library(data.table)
library(ggplot2)
library(lubridate)
```

## Introduction to GHCN-D and AWS

The Global Historical Climatology Network - Daily (GHCN-D) is a comprehensive database of daily climate summaries from land surface stations across the globe. This dataset is maintained by NOAA's National Centers for Environmental Information (NCEI) and includes records from over 100,000 stations in 180 countries and territories.

The GHCN-D dataset is now available on Amazon Web Services (AWS) S3, making it easier to access and analyze this valuable climate data. The `ghcn` package provides tools to efficiently work with this AWS-hosted dataset.

### Comparing AWS Bucket vs NOAA NCEI API

While the GHCN-D data has traditionally been accessible through the NOAA NCEI API, the availability on AWS offers some distinct advantages. Let's compare these two methods:

1. **Data Freshness**:
   - NOAA NCEI API: Provides the most up-to-date data, including recent observations.
   - AWS Bucket: Updated daily, but may have a slight lag compared to the API.

2. **Access Speed**:
   - NOAA NCEI API: Can be slower, especially for large data requests, due to server-side processing.
   - AWS Bucket: Generally faster, especially for bulk data access, as files are directly available.

3. **Query Flexibility**:
   - NOAA NCEI API: Offers more flexible querying options, allowing for complex filters on the server-side.
   - AWS Bucket: Requires client-side filtering, but the `ghcn` package simplifies this process.

4. **Data Format**:
   - NOAA NCEI API: Provides data in various formats, including CSV and JSON.
   - AWS Bucket: Data is in CSV format, which the `ghcn` package efficiently processes.

5. **Usage Limits**:
   - NOAA NCEI API: Has rate limits and may require authentication for large requests.
   - AWS Bucket: No rate limits for data access, but standard AWS data transfer costs may apply.

6. **Ease of Bulk Data Access**:
   - NOAA NCEI API: Can be cumbersome for accessing large volumes of data.
   - AWS Bucket: Excels at providing easy access to large datasets.

7. **Integration with Cloud Computing**:
   - NOAA NCEI API: Requires data to be downloaded to local or cloud storage before processing.
   - AWS Bucket: Allows for direct integration with AWS cloud computing services.

The `ghcn` package leverages the advantages of the AWS-hosted dataset, providing fast and efficient access to GHCN-D data without the need to worry about API rate limits or complex query constructions. This makes it particularly suitable for large-scale data analysis and research projects that require access to extensive historical climate data.

While the NOAA NCEI API remains valuable for accessing the most current data and for very specific queries, the AWS bucket approach used by the `ghcn` package offers a streamlined experience for most research and analysis needs, especially when working with historical data at scale.

## Installation

You can install the `ghcn` package using one of the following methods:

### Using remotes

The easiest way to install the package is using the `remotes` package:

```r
# install.packages("remotes")
remotes::install_github("joshbrinks/ghcn")
```

### Cloning the repository

If you want to clone the repository and install the package locally, follow these steps:

1. Clone the repository:
   ```
   git clone https://github.com/joshbrinks/ghcn.git
   ```

2. Open R or RStudio and set your working directory to the cloned repository:
   ```r
   setwd("path/to/ghcn")
   ```

3. Install the package:
   ```r
   devtools::install()
   ```

### Development version

To install the development version of the package, you can use:

```r
remotes::install_github("joshbrinks/ghcn@develop")
```

This will install the latest version from the `develop` branch.

After installation, you can load the package using:

```r
library(ghcn)
```

## Using the ghcn Package

The `ghcn` package offers three main functions to work with the GHCN-D data on AWS:

1. `find_stations_aws()`: Find weather stations near a given location.
2. `get_inventory_aws()`: Retrieve inventory information for specific stations.
3. `get_data_aws()`: Fetch actual weather data for specified stations and time periods.

Let's walk through how to use these functions together.

### Finding Nearby Stations

First, let's find weather stations near New York City:

```{r find_stations}
nyc_stations <- find_stations_aws(lat = 40.7128, lon = -74.0060, radius = 50)
head(nyc_stations)
```

This gives us a list of stations within a 50km radius of NYC, sorted by distance.

### Getting Station Inventory

Now, let's check what data is available for the first two stations:

```{r get_inventory}
station_ids <- nyc_stations$id[1:2]
inventory <- get_inventory_aws(station_ids)
head(inventory)
```

This inventory shows us what variables are available for each station and their period of record.

### Fetching Weather Data

Let's fetch temperature data for the first station for the year 2022:

```{r get_data}
station_data <- get_data_aws(
  station_id = station_ids[1],
  start_date = "2022-01-01",
  end_date = "2022-12-31",
  variables = c("TMAX", "TMIN")
)
head(station_data)
```

## Demonstrating Functionality

### Multiple Stations

We can fetch data for multiple stations at once:

```{r multiple_stations}
multi_station_data <- get_data_aws(
  station_id = station_ids,
  start_date = "2022-01-01",
  end_date = "2022-12-31",
  variables = c("TMAX", "TMIN", "PRCP")
)
head(multi_station_data)
```

### Long Format Data

By default, `get_data_aws()` returns data in wide format. We can get long format data by specifying the `format` parameter:

```{r long_format}
long_data <- get_data_aws(
  station_id = station_ids[1],
  start_date = "2022-01-01",
  end_date = "2022-12-31",
  variables = c("TMAX", "TMIN", "PRCP"),
  format = "long"
)
head(long_data)
```

## Data Visualization and Analysis

Let's visualize the temperature data for one station using ggplot2:

```{r visualize_temp}
ggplot(station_data, aes(x = DATE)) +
  geom_line(aes(y = TMAX, color = "Max Temp")) +
  geom_line(aes(y = TMIN, color = "Min Temp")) +
  scale_color_manual(values = c("Max Temp" = "red", "Min Temp" = "blue")) +
  labs(title = "Daily Temperature Range in 2022",
       subtitle = paste("Station:", station_ids[1]),
       x = "Date", y = "Temperature (tenths of degrees Celsius)",
       color = "Temperature Type") +
  theme_minimal()
```

We can also look at the distribution of daily temperature ranges:

```{r temp_range}
station_data[, temp_range := TMAX - TMIN]

ggplot(station_data, aes(x = temp_range)) +
  geom_histogram(binwidth = 10, fill = "skyblue", color = "black") +
  labs(title = "Distribution of Daily Temperature Ranges",
       subtitle = paste("Station:", station_ids[1]),
       x = "Temperature Range (tenths of degrees Celsius)",
       y = "Count") +
  theme_minimal()
```

Let's analyze precipitation patterns:

```{r precip_analysis}
multi_station_data[, month := month(DATE)]
monthly_precip <- multi_station_data[, .(total_precip = sum(PRCP, na.rm = TRUE)),
                                     by = .(ID, month)]

ggplot(monthly_precip, aes(x = factor(month), y = total_precip, fill = ID)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_x_discrete(labels = month.abb) +
  labs(title = "Monthly Precipitation in 2022",
       x = "Month", y = "Total Precipitation (tenths of mm)",
       fill = "Station ID") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Conclusion

The `ghcn` package provides powerful tools for accessing and analyzing GHCN-D data from AWS. By combining these functions with data manipulation and visualization libraries like data.table and ggplot2, you can easily explore climate patterns and trends.

Remember to cite the GHCN-D dataset in your work:

Menne, M.J., I. Durre, B. Korzeniewski, S. McNeal, K. Thomas, X. Yin, S. Anthony, R. Ray, R.S. Vose, B.E.Gleason, and T.G. Houston, 2012: Global Historical Climatology Network - Daily (GHCN-Daily), Version 3. [indicate subset used following decimal, e.g. Version 3.12]. NOAA National Climatic Data Center. http://doi.org/10.7289/V5D21VHZ [access date].
